---
title: "Livability Figures"
html:
  self-contained: true   
editor: visual
---

These are some figures for the livability paper

## Load packages

```{r, message=FALSE, results='hide', warning=FALSE}
#| code-fold: true

library(here)
library(tidyverse)
library(sf)
library(tigris)
library(rnaturalearth)
library(ggthemes)
library(GGally)

sf_use_s2(FALSE)
```

## Load data

City-level data assembled in data-assembly.

```{r, message=FALSE, results='hide'}
#| code-fold: true

city_data <- here("data-for-figs.geojson") |>
  st_read()
```

Create a basic basemap.

```{r, message=FALSE, results='hide'}
#| code-fold: true

states_map <- states() |>
  filter(STUSPS %in% c("RI", 
                       "NJ",
                       "CT",
                       "IL",
                       "IN",
                       "MA",
                       "ME",
                       "MN",
                       "MI",
                       "ND",
                       "NH",
                       "NY",
                       "OH",
                       "PA",
                       "VT",
                       "WI")) 

great_lakes <- ne_download(type = "lakes", category = "physical", scale = 50) |>
  st_transform(crs = st_crs(states_map)) |>
  st_filter(states_map)

cape_cod <- area_water(state = "MA", county = c("Barnstable", 
                                                "Plymouth", 
                                                "Suffolk",
                                                "Nantucket",
                                                "Dukes"))

long_island <- area_water(state = "NY", county = c("Suffolk", 
                                                   "Nassau",
                                                   "Queens",
                                                   "Kings",
                                                   "New York"))

nj_water <- area_water(state = "NJ", county = c("Bergen", 
                                                "Hudson",
                                                "Union",
                                                "Middlesex",
                                                "Monmouth"))

crs_frostbelt = "+proj=lcc +lat_0=39 +lon_0=-89 +lat_1=37 +lat_2=47 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +type=crs"

base_map <- ggplot(states_map) +
  geom_sf(fill = "gray90",
          color = "white") +
  geom_sf(data = great_lakes,
          color = NA, 
          fill = "white") +
  geom_sf(data = cape_cod, fill = "white", color = NA) +
  geom_sf(data = long_island, fill = "white", color = NA) +
  geom_sf(data = nj_water, fill = "white", color = NA)  +
  theme_map()

```

## Maps

Here is a map showing spatial variation in housing affordability.

```{r}
#| code-fold: true

base_map +
  geom_sf(data = city_data,
          color = NA,
          aes(fill = housing_trunc)) +
  scale_fill_viridis_c(name = "Median housing\ncosts as a share\nof median income",
                       breaks = seq(0.25, 0.5, by=0.05),
                       labels = c(paste0(seq(25, 45, by=5),"%"), "50% or more")) +
  coord_sf(crs = crs_frostbelt)


```

And unemployment.

```{r}
#| code-fold: true

base_map +
  geom_sf(data = city_data,
          color = NA,
          aes(fill = unemp_trunc)) +
  scale_fill_viridis_c(name = "Unemployment\nrate",
                       breaks = seq(0, 0.1, by = 0.02),
                       labels = c(paste0(seq(0, 8, by=2), "%"), "10% or more")) +
  coord_sf(crs = crs_frostbelt)
```

And transportation access.

```{r}
#| code-fold: true

base_map +
  geom_sf(data = city_data,
          color = NA,
          aes(fill = access_trunc)) +
  scale_fill_viridis_c(name = "Car-free\naccessibility\n(relative to\naccess by car",
                       breaks = seq(0.05, 0.35, by = 0.05),
                       labels = c("5% or less", 
                                  paste0(seq(10, 35, by=5), "%")),
                       direction = -1) +
  coord_sf(crs = crs_frostbelt)
```

And adaptation plan (or hazard mitigation plan) quality

```{r}
#| code-fold: true

base_map +
  geom_sf(data = city_data,
          color = NA,
          aes(fill = plan_pts)) +
  scale_fill_viridis_c(name = "Plan quality",
                       breaks = seq(0, 25, by = 5),
                       direction = -1) +
  coord_sf(crs = crs_frostbelt)
```

## Scatter plots

This one isn't necessarily for the publication, but just for a quick look.

```{r, message=FALSE, warning=FALSE}
#| code-fold: true

ggpairs(
  city_data,
  columns = c(2,4, 6, 8),          # Select numeric columns
  aes(size = populationE),   # Map color to a factor
  upper = list(continuous = wrap("cor", size = 4)), # Show correlation in upper panel
  lower = list(continuous = wrap("points", alpha = 0.6, size = 1.5))
)



```

### Comparing plan quality with transportation

```{r}
#| code-fold: true

ggplot(city_data) +
  geom_point(aes(x = plan_pts,
                 y = avg_access_ratio,
                 size = populationE),
             alpha = 0.4) +
  scale_x_continuous(name = "Plan quality score") +
  scale_y_continuous(
    name = "Ratio of car-free accessibilty\nto accessibilty by car") +
  scale_size(name = "Population",
             breaks = breaks <- c(2*10^seq(2:7)),
             labels = formatC(breaks, format = "d", big.mark = ",")) +
  theme_minimal()

```

### Comparing plan quality with housing

```{r}
#| code-fold: true

ggplot(city_data) +
  geom_point(aes(x = plan_pts,
                 y = housing_trunc,
                 size = populationE),
             alpha = 0.4) +
  scale_x_continuous(name = "Plan quality score") +
  scale_y_reverse(
    name = "Ratio of median housing costs\nto median income",
    breaks = seq(0.1, 0.5, by = 0.05),
    labels = c(seq(0.1, 0.45, by= 0.05), "0.5 or more")) +
  scale_size(name = "Population",
             breaks = breaks <- c(2*10^seq(2:7)),
             labels = formatC(breaks, format = "d", big.mark = ",")) +
  theme_minimal()
```

### Comparing plan quality to unemployement

```{r}
#| code-fold: true

ggplot(city_data) +
  geom_point(aes(x = plan_pts,
                 y = unemp_trunc,
                 size = populationE),
             alpha = 0.4) +
  scale_x_continuous(name = "Plan quality score") +
  scale_y_reverse(
    name = "Unemployment rate",
    breaks = seq(0, 0.1, by = 0.02),
    labels = c(paste0(seq(0, 8, by= 2), "%"), "10% or more")) +
  scale_size(name = "Population",
             breaks = breaks <- c(2*10^seq(2:7)),
             labels = formatC(breaks, format = "d", big.mark = ",")) +
  theme_minimal()
```
